Building Trust in Science Through a Bayesian Perspective on Probability and Uncertainty

### 2021-04-29 manuscript.tex
changes from e.j.'s adds

\usepackage[american]{babel}
\usepackage{nicefrac}
\usepackage{todonotes}
\usepackage{epigraph}
% For sorting references from same year in order instead of on title
\newcommand{\SR}[1]{} %for EJ's bibtex file
\newcommand{\EJ}[1]{\todo[inline, color=green]{  #1 }}
\newcommand{\JR}[1]{\todo[inline, color=gray]{  #1 }}
\newcommand{\MK}[1]{\todo[inline, color=red]{  #1 }}

@@

\usepackage{csquotes}
\usepackage[style=apa,sortcites=true,sorting=nyt,backend=biber]{biblatex}
\DeclareLanguageMapping{american}{american-apa}
\addbibresource{bibliography.bib}

\title{Trust Me, I'm a Bayesian: Building Back Trust in Science Using Bayes}
\shorttitle{Bayes in Science Education}

\author{Joshua M. Rosenberg}
\affiliation{University of Tennessee, Knoxville}

\author{Marcus Kubsch}
\affiliation{IPN - Leibniz Institute for Science and Mathematics Education}

\author{Eric-Jan Wagenmakers}
\affiliation{University of Amsterdam}

@@

\abstract{

Uncertainty is ubiquitous in science and this uncertainty takes many forms, but the rhetoric around scientific knowledge and negotiating standards of evidence is often portrayed as immutable. Adding to this problem around the limited recognition of the tentative nature of science, when scientists try to communicate their uncertainty, research has shown us that many—including both learners and experts—struggle to interpret the statistics that are often used to make inferences from scientific data. We argue that a Bayesian approach can offer a different foundation and language around a core but sometimes unsettling tenet and persistent feature, uncertainty. Central to this conceptualization is viewing uncertainty probabilistically, as probabilities can provide a language for expressing the degree of uncertainty in our knowledge in a wide range of domains. In this essay, we a) briefly introduce Bayes theorem and wider ideas around Bayesian epistemology, b) elaborate on the importance of probability and uncertainty in science and science education and review how people learn about these concepts and where they face difficulties, and c) provide conceptual and practical examples within the JASP statistical software of how Bayes could help to address these issues. We conclude with implications and directions for future research focused around bolstering trust in science outside of classroom settings (for which Bayesian approaches could serve as a crosscutting part of science literacy) and in pre-collegiate science classrooms, where Bayes can provide a coherent approach to engaging in numerous science practices such as arguing from evidence and analyzing and interpreting data in a principled and flexible way.
}

@@

\authornote{
   \addORCIDlink{Joshua M. Rosenberg}{0000-0003-2170-0447}
   \addORCIDlink{Marcus Kubsch}{0000-0000-0000-0000}
   \addORCIDlink{Eric-Jan Wagenmakers}{0000-0003-1596-1034} 
}

@@

\EJ{This is a comment.}

Uncertainty is ubiquitous in science and this uncertainty takes many forms. For example, measurement uncertainty is present when there are differences (small or large) in the repeated measures of the same object or phenomenon \parencite{pls03} or when the object that is measured, like a plant, is inherently variable in its characteristics \parencite{pls03}. Another form of uncertainty is its role within explanatory theories, such as those associated with evolution and quantum physics, for which central models take the form of probablistic statments, i.e., uncertainty is inherent in these models (Gigerenzer, 2000).

Negotiating these many forms of uncertainty in argumentation—and building toward consensus at the same time that uncertainty is present—is a key part of the scientific process (Thagard, 2000). In consequence, some philosophers of science have suggested that science is a process that builds increasingly better models which allows us to make increasingly more accurate predictions about the world (Giere, 2010; Godfrey-Smith, 2006; Nersessian, 2002; Reichenbach, 1977).

However, while scientists and philosophers of science may be aware of the uncertainty (and limitations) surrounding scientific knowledge and negotiating standards of evidence interpretation, scientific knowledge is still often portrayed as immutable in educational contexts (Duschl et al., 1990). In consequence, trust in science can erode when scientists individually or collectively change their views, such as during the first phase of the COVID-19 crisis in Germany, when scientists faced serious criticism from politicians and the media for changing their opinion based on new information. A similar tension was present in the United States and around the world regarding potential COVID-19 medication, when initial scientific data and changing interpretations of its utility collided with citizens’ skepticism (and, partisan political views).

Adding to this problem around the limited recognition of the tentative nature of science, when scientists try to communicate their uncertainty, research has shown us that many—including learners and experts alike—struggle to interpret statistics such as confidence intervals or error bars (Gigerenzer et al. 2004; Sedlmeier, 2007; Tversky & Kahneman, 1977). Similarly, difficulties in understanding the language of probability can contribute to a struggle on the part of students when learning about concepts such as evolution of nuclear decay that are modelled in terms of probabilistic statements (e.g., Fiedler et al., 2019).

Based on accumulating evidence from cognitive psychology, neuroscience, and education, we build on and extend beyond the science and engineering practice of argumentation the foundational work of science education (e.g., Nussbaum, 2011; Szu & Osborne, 2012) and statistics education scholars (e.g., Albert, 2002; Bolstad, 2002; Hoegh et al., 2020; Hu, 2020; Martignon and Erickson, 2014, Sedlmeier, 2007; Witmer, 2017) to argue for a relatively novel approach to uncertainty and probability. Specifically, we argue that it is helpful to consider uncertainty in terms of a Bayesian epistemological framework to help people to understand uncertainty and probability better and to engage with them in a way that leverages their existing ways of thinking. Central to this conceptualization is viewing uncertainty probabilistically, as probabilities can provide a language for expressing the degree of uncertainty in our knowledge in a wide range of domains (Gopnik, 2012). In this way, we make this conjecture not only on the merits of Bayes Theorem as a mathematical or statistical tool, but based on a body of psychological and educational research that shows that such an approach is an intuitive way for people to think about data and evidence in the world (Gopnik, 2012) and in classroom contexts (Leuders & Loibl, 2020; Nussbaum, 2011; Sedlmeier, 2007; Szu & Osborne, 2012; Witmer, 2017).
The remainder of this paper is structured in the following way. We will first briefly introduce Bayes theorem and wider ideas around Bayesian epistemology. Next, we will elaborate on the importance of probability and uncertainty in science (education) and review how people learn about these concepts and where they face difficulties. Then, we will provide examples of how Bayes could help to address these issues before concluding with implications and directions for future research.

@@

In essence, Bayes’ theorem provides a framework for updating one’s beliefs about the world from observations in a way that takes prior knowledge into account in a principled manner. Bayes theorem (Equation 1) states that our knowledge about the world after making an observation can be expressed as a probability distribution over the possible outcomes of the phenomenon we observed—the posterior distribution. To calculate the posterior, we express our knowledge about the world before the observation in terms of a probability distribution, the prior, and multiply that with the likelihood which summarizes the information in our observation. These are then divided by the average likelihood to yield a posterior distribution that sums to one.:

Posterior = Prior x LikelihoodAverage Likelihood

\EJ{So I thought about introducing this more elaborately with an example where interest centers on what proportion of students is using
Windows vs. the Mac. Data from the first 10 students are $\{W, W, M, W, M, W, W, M, W, M\}$. I'm using a beta$(2,2)$ prior. The results are shown in the two figures below.}

These are big ideas, and, fortunately, there are a number of intuitive explanations of Bayesian methods, including Kurt (2019), Kruschke and Liddell (2018a) Wagenmakers et al. (2016). A recent example from physics can serve to illustrate this principle in the context of science education. In 2011, physicists reported that they had observed elementary particles moving faster than the speed of light (Brumfiel, 2011) - and thus defying Einstein’s theory of relativity. However, instead of celebrating a historic discovery, the scientists started to look for what had gone wrong in their observations.

@@

Thus, rather than changing their belief about the speed of light, scientists looked for things that could have gone wrong with the measurement. This example shows how a qualitative take on Bayes’ theorem can explain and predict many aspects of how animals (Okasha et al., 2012), people (Griffiths et al. 2008, Tenenbaum et al. 2006), and organisations (Kruschke et al., 2012) learn from observations. Recently, advances in computational power and techniques have started the more and more widespread application of this principle in a quantitative way that allows researchers to precisely express their knowledge about the world and then learn from observations.

Expressed in the Bayesian learning cycle (Figure 1), a Bayesian reasoning framework allows scientists to integrate the strengths of both, inductive and deductive research approaches, in a formalized way. Further, when specifying the prior, scientists openly and explicitly need to describe the (un)certainty in the current state of knowledge and the posterior captures how the data has changed the knowledge and the (un)certainty in that knowledge. Thus, Bayesian reasoning is, at least in principle, open and transparent by design.

Figure 1
The Bayesian learning cycle adapted from Wagenmakers and Gronau (2018).
\includegraphics[scale = .75]{learning.png}

@@

Probability and uncertainty are ubiquitous in science—in scientific methodologies, the scientific concepts, or in how science is communicated (Gougis et al., 2016). The practice of science often begins with measurement, and with measurement comes measurement error, which has a random component. This random component arises from random variations in the measurement process and limits the certainty that one can have in the measurements. Reducing that measurement error has been critical for many discoveries in science, such as in sorting out the periodic table as we know it today (Fontani, 2014). However, deciding on adequate statistical procedures to determine measurement uncertainty regularly sparks debate, e.g., the recent study on the relationship between SARS-CoV-2 viral load and patient age by Jones et al. (2020) was heavily debated in the scientific community (Frick, 2020).

Probability and uncertainty are not only key concepts in science practices such as analyzing and interpreting data, but also part of the scientific concepts themselves. One example is Heisenberg’s uncertainty principle or quantum mechanics more generally  (Feynman, 1951). In quantum mechanics, confidence in measurement is limited not only by measurement error but in principle—even in the perfect experimental setup there is no deterministic outcome. Instead, one has to calculate the probability of each possible outcome using the Born rule. To this day, there is an ongoing argument among physicists and philosophers about how to interpret probability and the puzzling quantum mechanical phenomena such as quantum tunnelling (e.g., Caroll [2020]). Another example of how probability and uncertainty are central to learning scientific concepts is the probabilistic nature of evolution (Fiedler et al., 2017). Evolution is a probabilistic process as the emergence of new variations (e.g., through mutations) can only be described probabilistically. Further, the continued survival of these variations is governed by phenomena such as natural selection or random drift which also escape a deterministic description and are thus modelled probabilistically. Similar to within quantum physics, how to interpret this probability has also sparked debate among researchers (Millstein, 2014).

Given that probability and uncertainty spark debate among scientists, it is perhaps unsurprising that misinterpretations happen when scientists in the natural and engineering sciences communicate about their findings with their peers or the public (e.g., Gigerenzer 2004, McShane & Gal, 2017, Cumming, 2014). For example, error bars in the form of confidence intervals are routinely interpreted as distributions that assign higher probability to the center of the interval (Kruschke & Liddell, 2018a) or p-values are interpreted as effect sizes (Gelman & Carlin, 2017; Nuzzo; 2014) although p-values describe how (in)compatible a set of data is with a set of assumptions. In sum, misinterpretations of statements about probability and uncertainty are common across the range of modes in which scientists communicate about their work.

@@

Given the above-noted roles of probability in scientific concepts and theories, it is unsurprising that educational researchers have documented how students learn about probability. As one example, probability and chance events play an important role in the life sciences (Garfield, 2003; Garvin-Doxas & Klymkowski, 2017), particularly in evolutionary processes and learning about them (Tibell & Harms, 2017). Recently, Fiedler et al. (2019) investigated the relation between statistical reasoning and acceptance and knowledge about evolution in a large sample of nearly 500 university students in the US. They found that students’ statistical reasoning capabilities are a strong predictor of both acceptance of evolution and knowledge about evolution. Thus, this study suggests (but does not directly show) that instructional strategies for supporting students’ statistical reasoning could improve their learning about specific evolutariony concepts.

In the physical sciences, students’ struggles with the probabilistic nature of quantum physics (Bao & Redish, 2002) and nuclear decay (Santostasi et al., 2017) are known, but the underlying reasons and mechanisms remain less researched than in evolution. Research into students’ misconceptions inspired by conceptual change research  (e.g., Marshman & Singh, 2015, Aguilar et al., 2014, Stefani & Tsarparlis, 2009) has started to address this gap. Findings seem to mirror the position of Batanero et al. (1994) in that the concepts of probability are taught too abstractly and in a way that does not build upon and connect with the ideas that students’ already hold, the experiences that students make, and the language that students use. Thus, these studies affirm students' struggles with scientific concepts that are probabilistic but hint at the underlying issue being foundational, conceptual struggles with ideas about probability not limited to science learning. Indeed, the language and representations used to communicate probability and uncertainty independent of specific scientific content domains or scientific practices are also known to present difficulties to learners. The work of Tversky and Kahneman (1974) points to a whole range of biases that the language in which statements about probability and uncertainty are framed can elicit. A prominent example is the neglect of base rate information in the (timely) context of medical testing, that is, the probability of having a medical condition on the basis of a positive test result is often hugely overestimated because information about the base rate of people affected from the condition is neglected (Kahneman, 2012).

More recent work by Gigerenzer and colleagues has provided important insights into the underlying mechanisms for the occurence of these biases (e.g., Gigerenzer & Hoffrange, 1995; Jenny et al., 2018). Their research suggests that the biases are at least partly the result of information about probability and uncertainty being provided in a way that is incompatible with the heuristics that people develop from their everyday experiences. Based on this assumption, Jenny et al. (2018) were able to demonstrate that students judge probability significantly better, when language is adopted that better aligns with the heuristics that most learners develop from their everyday experiences. In sum, changing representations of probability and uncertainty so that they better align with ideas and heuristics that learners hold so that these support instead of inhibit learning seems feasible.

@@

Recent work in cognitive and psychological science on how children learn argues that children understand and learn about the world through the intuitive application of Bayesian ideas (Gopnik, 2012; Tenenbaum et al., 2006; Griffiths et al., 2008). This idea has led current research to conceptualize human learning in terms of Bayesian probabilistic models of learning. In other words, people (and children) may view (or come to view) the world probabilistically, and in ways that happen to align with Bayesian interpretations of probability.

This view of how children learn from data has become especially prominent in developmental science and is successfully applied there (Gopnik and Tenenbaum, 2007; Gopnik & Wellman, 2012). One might assume that there has been a ready application of these ideas about how children think about the world in a Bayesian way to education. Developmentalists have pointed to the potential and opportunity here. Gopnik (2012) wrote that the use of Bayesian methods to understand child development can serve as “a scientific foundation for a long tradition on ‘inquiry-based’ science education,” that “could lead us to much more specific and scientifically supported proposals for education” (p. 1627).

Despite arguments for integrating developmental and educational research, it has largely not been the case that—with respect to advances in Bayesian methods—“science itself could help turn young children’s natural curiosity and brilliance into better science teaching and learning” (Gopnik, 2012, p. 1627), as some science education scholars have bemoaned (Lehrer & Schauble, 2006). Indeed, scholars have argued that it is often obsolete applications of Piaget’s ideas (Piaget & Inhelder, 1969) that represent the greatest contribution of developmental ideas to education (Lehrer & Schauble, 2007; Metz, 1995). We argue that this is still largely the case in the present, despite advances in both developmental research as well as statistical methodology related to Bayesian ideas—and that it is time to start exploring how bringing out the inherent Bayesian ways of thinking in education may provide support for learners’ efforts to make sense of the world.

>>>>

\usepackage[american]{babel}
\usepackage{nicefrac}
\usepackage{todonotes}
\usepackage{epigraph}
\usepackage{amsfonts, amsmath,amssymb}

\newcommand{\SR}[1]{} %for EJ's bibtex file
\newcommand{\given}{\, | \,}
\newcommand{\EJ}[1]{\todo[inline, color=green]{  #1 }}
\newcommand{\JR}[1]{\todo[inline, color=gray]{  #1 }}
\newcommand{\MK}[1]{\todo[inline, color=red]{  #1 }}

@@

\usepackage{csquotes}
\usepackage[style=apa,sortcites=true,sorting=nyt,backend=biber]{biblatex}
\DeclareLanguageMapping{american}{american-apa}
\addbibresource{bibliography.bib}
\addbibresource{referenties.bib}

% \setlength{\parindent}{8ex}

% \usepackage{authblk}

\title{Building Trust in Science Through a Bayesian Perspective on Probability and Uncertainty}
\shorttitle{Building Trust Through a Bayesian Perspective}

\threeauthors{Joshua M. Rosenberg}{Marcus Kubsch}{Eric-Jan Wagenmakers}
\threeaffiliations{University of Tennessee, Knoxville}{IPN - Leibniz Institute for Science and Mathematics Education}{University of Amsterdam}

% \authorsnames[1,2,3]{Joshua M. Rosenberg, Marcus Kubsch, Eric-Jan Wagenmakers}
% \authorsaffiliations{{University of Tennessee, Knoxville},{IPN - Leibniz Institute for Science and Mathematics Education},{University of Amsterdam}}

@@

\abstract{
Uncertainty is ubiquitous in science and this uncertainty takes many forms, but the rhetoric around scientific knowledge and negotiating standards of evidence is often portrayed as immutable. Adding to this problem around the limited recognition of the tentative nature of science, when scientists try to communicate their uncertainty, research has shown us that many—including both learners and experts—struggle to interpret the statistics that are often used to make inferences from scientific data. We argue that a Bayesian approach can offer a different foundation and language around a core but sometimes unsettling tenet and persistent feature, uncertainty. Central to this conceptualization is viewing uncertainty probabilistically, as probabilities can provide a language for expressing the degree of uncertainty in our knowledge in a wide range of domains. In this essay, we (a) briefly introduce Bayes' theorem and wider ideas around Bayesian epistemology, (b) elaborate on the importance of probability and uncertainty in science and science education and review how people learn about these concepts and where they face difficulties, and (c) provide conceptual and practical examples within the JASP statistical software of how Bayes could help to address these issues. We conclude with implications and directions for future research focused around bolstering trust in science outside of classroom settings (for which Bayesian approaches could serve as a crosscutting part of science literacy) and in pre-collegiate science classrooms, where Bayes can provide a coherent approach to engaging in numerous science practices such as arguing from evidence and analyzing and interpreting data in a principled and flexible way.
}

@@

\authornote{
   \addORCIDlink{Joshua M. Rosenberg}{0000-0003-2170-0447}
   \addORCIDlink{Marcus Kubsch}{0000-0001-5497-8336}
   \addORCIDlink{Eric-Jan Wagenmakers}{0000-0003-1596-1034} 
   \\
   Joshua M. Rosenberg and Marcus Kubsch contributed equally to this manuscript.
}

@@

\EJ{I have doubts about the ``Trust me, I'm a Bayesian'' part of the title, also because ``Trust'' now occurs twice. Maybe simply ``Building Trust in Science Through a Bayesian Perspective on Probability and Uncertainty''?}
\MK{Ok with the new title - I liked the sound of the trust me thing but utlimatley it may not be really on point!}
\EJ{I also uploaded my bibtex file, ``referenties.bib''. I could not get it to work, but you could copy the relevant references into your own bibtex file. Actually, I also managed to make the initial bibtex setup not work, and I have absolutely no clue as to how this happened.}

Uncertainty is ubiquitous in science and this uncertainty takes many forms. For example, measurement uncertainty is present when there are differences (small or large) in the repeated measures of the same object or phenomenon \parencite{pls03} or when the object that is measured, like a plant, is inherently variable in its characteristics \parencite{pls03}. Another form of uncertainty is its role within explanatory theories, such as those associated with evolution and quantum physics, for which central models take the form of probablistic statements, i.e., uncertainty is a core constituent of these models \parencite{g00}.

\EJ{Yeah so as a hard-nosed determinist I still have an issue with ``inherent'' uncertainty. If we don't sort this out we could resolve it through a footnote (e.g., ``EJ Wagenmakers would like to note that he is a determinist who believes that the concepts of probability and uncertainty belongs wholly to the mind \parencite{Jevons18741913}, with the possible exception of the strange world of quantum mechanics.''} 
\MK{I got rid of the inherent part.}

\EJ{It would be good to go over the reference list in detail; in particular, I noticed several ``bayes'' that should be ``Bayes'' (the B needs to be enclosed in brackets so it displays in upper case)}
Negotiating these many forms of uncertainty in argumentation—and building toward consensus at the same time that uncertainty is present—is a key part of the scientific process \parencite{t00}. In consequence, some philosophers of science have suggested that science is a process that builds increasingly better models which allows us to make increasingly more accurate predictions about the world  \parencite{g10, g06, n02, r77}.

However, while scientists and philosophers of science may be aware of the uncertainty (and limitations) surrounding scientific knowledge and negotiating standards of evidence interpretation, scientific knowledge is still often portrayed as immutable in educational contexts \parencite{d90}. In consequence, trust in science can erode when scientists individually or collectively change their views, such as during the first phase of the COVID-19 crisis in Germany, when scientists faced serious criticism from politicians and the media for changing their opinion based on new information. A similar tension was present in the United States and around the world regarding potential COVID-19 medication, when initial scientific data and changing interpretations of its utility collided with citizens’ skepticism (and, partisan political views).

Adding to this problem around the limited recognition of the tentative nature of science, when scientists try to communicate their uncertainty, research has shown us that many—including learners and experts alike—struggle to interpret statistics such as confidence intervals or error bars \parencite{gkv04, s07, tk74}. Similarly, difficulties in understanding the language of probability can contribute to a struggle on the part of students when learning about concepts such as evolution of nuclear decay that are modelled in terms of probabilistic statements \parencite{fth17}.

Based on accumulating evidence from cognitive psychology, neuroscience, and education, we build on and extend beyond the science and engineering practice of argumentation the foundational work of science educational \parencite{n11, s07} and statistics education scholars \parencite{a02, b02, h_a20, h_j20,me14, s07} to argue for a \EJ{I'd say ``underused'', because Bayes itself is pretty old and now pretty hip} approach to uncertainty and probability that is currently underrepresented in science education (research). Specifically, we argue that it is helpful to consider uncertainty in terms of a Bayesian epistemological framework to help people to understand uncertainty and probability better and to engage with them in a way that leverages their existing ways of thinking. Central to this conceptualization is viewing uncertainty probabilistically, as probabilities can provide a language for expressing the degree of uncertainty in our knowledge in a wide range of domains \parencite{g12}. In this way, we make this conjecture not only on the merits of Bayes' Theorem as a mathematical or statistical tool, but based on a body of psychological and educational research that shows that such an approach is an intuitive way for people to think about data and evidence in the world \parencite{g12} and in classroom contexts \parencite{ll20, n11, s07, so12, w17}.

The remainder of this paper is structured in the following way. We will first briefly introduce Bayes theorem and wider ideas around Bayesian epistemology. Next, we will elaborate on the importance of probability and uncertainty in science (education) and review how people learn about these concepts and where they face difficulties. Then, we will provide examples of how Bayes could help to address these issues before concluding with implications and directions for future research.

@@

In essence, Bayes' theorem provides a framework for updating one's beliefs about the world from observations in a way that takes prior knowledge into account in a principled manner. Bayes' theorem (Equation 1) states that at any point in time, our knowledge about the world after making a new observation (i.e., our posterior knowledge) can be obtained by multiplying our prior knowledge with a predictive updating factor (e.g., \cite{RouderMorey2019,WagenmakersEtAl2016CD} ):
\begin{equation*}
        \text{Posterior uncertainty} = \text{Prior uncertainty} \times \text{Predictive updating factor}.
\end{equation*}
Mathematically, this can be expressed as  
\begin{equation}
\label{eq:BayesRule}
        p(\theta \mid \text{data}) = p(\theta) \times \frac{p(\text{data} \mid \theta)}{p(\text{data})},
\end{equation}
where $\theta$ is a hypothesis, proposition, claim, parameter value, or in general any particular account of the world about which we are uncertain. The predictive updating factor consists of two components. First, $p(\text{data} \given \theta)$ is the likelihood, that is, the extent to which the observed data were expected under $\theta$. Second, $p(\text{data})$ is the marginal likelihood, that is, the extent to which the observed data were expected without conditioning on $\theta$. If the observed data are more expected (i.e., better predicted) when conditioning on $\theta$, the predictive updating factor is higher than 1, and $\theta$ becomes more credible.

To make this more tangible, consider the following example. Suppose you are interested in the proportion of students that use Windows vs. MacOS as their operating system. The proportion of Windows users $\theta$ ranges between 0 and 1. Before you start our Bayesian analysis, you need to quantify your background knowledge -- the relative plausibility of the different values for $\theta$. Given your previous experiences in class rooms your best guess is that about half of the students use MacOS. You would not be surprised if the  proportion of Windows users was between 30\% and 70\% but you would be rather surprised to find either no Windows or exclusive Windows usage. This background knowledge motivates the use of the dome-shaped prior distribution shown as the dashed line in Figure~\ref{fig:WindowsMacPriorPosterior}.\footnote{This is the beta$(2,2)$ distribution.} Now suppose you start to collect data by asking students about their operating systems. The first ten observations of Windows (W) and MacOS (M) usage that you make are $\{W, W, M, W, M, W, W, M, W, M\}$: six students use Windows, four use MacOS. With these data in hand, we then combine our prior knowledge (i.e., the dome-shaped prior distribution) with the predictive updating factor (assuming a binomial likelihood) and obtain the posterior distribution shown as the solid line in Figure~\ref{fig:WindowsMacPriorPosterior}. The posterior distribution is more peaked than the prior distribution, indicating that the data have sharpened our knowledge about $\theta$. In addition, the posterior distribution is higher than the prior distribution for values of $\theta$ between approximately $.4$ and $.75$: these values of $\theta$ predicted the data relatively well and have therefore gained credibility. In contrast, values of $\theta$ lower than $.4$ and higher than $.75$ predicted the data relatively poorly, which is why they have lost credibility compared to the prior.\footnote{Note that the area under a probability distribution equals 1, so that posterior gains for specific regions of $\theta$ cancel out exactly against posterior losses for the other regions.}

\begin{figure*}[h]
\begin{center}
\includegraphics[width = .785\paperwidth]{WindowsMacPriorPosterior.pdf}
\caption{Simple example of Bayesian learning. A dome-shaped prior distribution captures the background knowledge concerning the proportion of students using Windows vs. MacOS. Observing fictitious data from 10 students (six using Windows, four using MacOS) drives a knowledge update that results in a bell-shaped posterior distribution. Figure based on the \emph{Learn Bayes} module in JASP.}
\label{fig:WindowsMacPriorPosterior}
\end{center}
\end{figure*}

Although we have demonstrated this updating process in a single step, it could also be executed sequentially, one student at a time. For instance, the first observation you make is a $W$ and in light of your prior this makes Windows usage a little more likely than MacOS usage. This slight change in knowledge is reflected in the difference between the distributions on the top two lines in Figure~\ref{fig:WindowsMacSequential}; line ``0'' represents the dome-shaped prior distribution and line ``1'' represents the posterior distribution after the first observations. Note that observing a $W$ has nudged the distribution a little to the right. Crucially, the `nudged' posterior distribution after the first observation now takes the role of the prior distribution ready to be updated by our second observation. The second observation is again a $W$, and line ``2'' shows that the resulting posterior distribution is nudged to the right once more. At every stage in the sequential updating process, the posterior distribution based on the observations seen so far becomes the prior distribution for incorporating the information from the next observation. Specifically, the changing distributions in Figure~\ref{fig:WindowsMacSequential} show our knowledge about the world is subject to constant change, by making observations and integrating incoming information with current knowledge using Bayes' theorem. It should be emphasized that although the updating process is conceptually different, the final outcome is identical: the shape of the posterior distribution is unaffected by whether data arrive simultaneously or sequentially.

\begin{figure*}[h]
\begin{center}
\includegraphics[width = .65\paperwidth]{WindowsMacSequential.pdf}
\caption{Sequential Bayesian learning. A dome-shaped prior distribution (line ``0'') captures the background knowledge concerning the proportion of students using Windows vs. MacOS. Each new observation results in an update to a posterior distribution, which then becomes the prior distribution for the analysis of the next observation. Figure based on the \emph{Learn Bayes} module in JASP.}
\label{fig:WindowsMacSequential}
\end{center}
\end{figure*}

But how do the principles from this example translate to science (education)? A recent case from physics can serve to illustrate this. In 2011, physicists reported that they had observed elementary particles moving faster than the speed of light \parencite{b11} - and thus defying Einstein’s theory of relativity. However, instead of celebrating a historic discovery, the scientists started to look for what had gone wrong in their observations.

@@

Thus, rather than changing their belief about the speed of light, scientists looked for things that could have gone wrong with the measurement. This example shows how a qualitative take on Bayes’ theorem can explain and predict many aspects of how animals \parencite{o13}, people \parencite{tgk06}, and organisations \parencite{kaj12} learn from observations. Recently, advances in computational power and techniques have started the more and more widespread application of this principle in a quantitative way that allows researchers to precisely express their knowledge about the world and then learn from observations.

Expressed in the Bayesian learning cycle (Figure \ref{fig:The Bayesian Learning Cycle}), a Bayesian reasoning framework allows scientists to integrate the strengths of both, inductive and deductive research approaches, in a formalized way. Further, when specifying the prior, scientists openly and explicitly need to describe the (un)certainty in the current state of knowledge and the posterior captures how the data has changed the knowledge and the (un)certainty in that knowledge. Thus, Bayesian reasoning is, at least in principle, open and transparent by design.

\begin{figure*}[h]
\begin{center}
\includegraphics[width = .65\paperwidth]{learning.png}
\caption{The Bayesian Learning cycle}
\label{fig:The Bayesian Learning Cycle}
\end{center}
\end{figure*}

@@

Probability and uncertainty are ubiquitous in science—in scientific methodologies, the scientific concepts, or in how science is communicated \textcite{gsoobmc17}. The practice of science often begins with measurement, and with measurement comes measurement error, which has a random component. This random component arises from random variations in the measurement process and limits the certainty that one can have in the measurements. Reducing that measurement error has been critical for many discoveries in science, such as in sorting out the periodic table as we know it today \textcite{fco15}. However, deciding on adequate statistical procedures to determine measurement uncertainty regularly sparks debate, e.g., the recent study on the relationship between SARS-CoV-2 viral load and patient age by \textcite{jmvbzhd20} was heavily debated in the scientific community \parencite{frick_peer-review_2020}.

Probability and uncertainty are not only key concepts in science practices such as analyzing and interpreting data, but also part of the scientific concepts themselves. One example is Heisenberg’s uncertainty principle or quantum mechanics more generally \textcite{feynman1951operator}. In quantum mechanics, confidence in measurement is limited not only by measurement error but in principle—even in the perfect experimental setup there is no deterministic outcome. Instead, one has to calculate the probability of each possible outcome using the Born rule. To this day, there is an ongoing argument among physicists and philosophers about how to interpret probability and the puzzling quantum mechanical phenomena such as quantum tunnelling  \textcite{c19}. Another example of how probability and uncertainty are central to learning scientific concepts is  evolution  \textcite{fsnh19}. Evolution is modelled as a probabilistic process as the emergence of new variations (e.g., through mutations) can only be described probabilistically. Further, the continued survival of these variations is governed by phenomena such as natural selection or random drift which also escape a deterministic description and are thus modelled probabilistically.\EJ{I agree that we can only describe the process probabilistically. But that does not mean it \emph{is} probabilistic.} \MK{tried to address this comment as well.} Similar to within quantum physics, how to interpret this probability has also sparked debate among researchers  \textcite{m16}.
Given that probability and uncertainty spark debate among scientists, it is perhaps unsurprising that misinterpretations happen when scientists in the natural and engineering sciences communicate about their findings with their peers or the public \parencite[]{gkv04, c14, mg17}. For example, error bars in the form of confidence intervals are routinely interpreted as distributions that assign higher probability to the center of the interval \parencite{kl18} or \emph{p}-values are interpreted as effect sizes \parencite{gc17, n14} (although p-values describe how (in)compatible a set of data is with a set of assumptions. In sum, misinterpretations of statements about probability and uncertainty are common across the range of modes in which scientists communicate about their work.

@@

Given the above-noted roles of probability in scientific concepts and theories, it is unsurprising that educational researchers have documented how students learn about probability. As one example, probability and chance events play an important role in the life sciences \parencite{g03, gk08}, particularly in evolutionary processes and learning about them \parencite{th17}. Recently, \textcite{fsnh19} investigated the relation between statistical reasoning and acceptance and knowledge about evolution in a large sample of nearly 500 university students in the US. They found that students’ statistical reasoning capabilities are a strong predictor of both acceptance of evolution and knowledge about evolution. Thus, this study suggests (but does not directly show) that instructional strategies for supporting students’ statistical reasoning could improve their learning about specific evolutionary concepts.

In the physical sciences, students' struggles with the probabilistic nature of quantum physics \parencite{br02} and nuclear decay \parencite{smmv17} are known, but the underlying reasons and mechanisms remain less researched than in evolution. Research into students’ misconceptions inspired by conceptual change research has started to address this gap \parencite{ms15, alra14, st09}. Findings seem to mirror the position of \textcite{bgv94} in that the concepts of probability are taught too abstractly and in a way that does not build upon and connect with the ideas that students’ already hold, the experiences that students make, and the language that students use. Thus, these studies affirm students' struggles with scientific concepts that are probabilistic but hint at the underlying issue being foundational, conceptual struggles with ideas about probability not limited to science learning. Indeed, the language and representations used to communicate probability and uncertainty independent of specific scientific content domains or scientific practices are also known to present difficulties to learners. The work of \textcite{tk74} points to a whole range of biases that the language in which statements about probability and uncertainty are framed can elicit. A prominent example is the neglect of base rate information in the (timely) context of medical testing, that is, the probability of having a medical condition on the basis of a positive test result is often hugely overestimated because information about the base rate of people affected from the condition is neglected \parencite{kahneman_thinking_2012}.

More recent work by Gigerenzer and colleagues has provided important insights into the underlying mechanisms for the occurrence of these biases \parencite{gh95, jkg18}. Their research suggests that the biases are at least partly the result of information about probability and uncertainty being provided in a way that is incompatible with the heuristics that people develop from their everyday experiences. Based on this assumption, \textcite{jkg18} were able to demonstrate that students judge probability significantly better, when language is adopted that better aligns with the heuristics that most learners develop from their everyday experiences. In sum, it seems feasible to change representations of probability and uncertainty so that they better align with ideas and heuristics that learners hold so that these support instead of inhibit learning.

@@

Recent work in cognitive and psychological science on how children learn argues that children understand and learn about the world through the intuitive application of Bayesian ideas \parencite{g12, tgk06, tgk06}. This idea has led current research to conceptualize human learning in terms of Bayesian probabilistic models of learning. In other words, people (and children) may view (or come to view) the world probabilistically, and in ways that happen to align with Bayesian interpretations of probability.

This view of how children learn from data has become especially prominent in developmental science and is successfully applied there \parencite{gt07, gw12}. One might assume that there has been a ready application of these ideas about how children think about the world in a Bayesian way to education. Developmentalists have pointed to the potential and opportunity here. \textcite{g12} wrote that the use of Bayesian methods to understand child development can serve as “a scientific foundation for a long tradition on ‘inquiry-based’ science education,” that “could lead us to much more specific and scientifically supported proposals for education” (p. 1627).

Despite arguments for integrating developmental and educational research, it has largely not been the case that—with respect to advances in Bayesian methods—“science itself could help turn young children’s natural curiosity and brilliance into better science teaching and learning” \textcite[p. 1627]{g12}, as some science education scholars have bemoaned \parencite{ls15}). Indeed, scholars have argued that it is often obsolete applications of Piaget’s ideas \textcite{pi69} that represent the greatest contribution of developmental ideas to education \textcite{ls07}. We argue that this is still largely the case in the present, despite advances in both developmental research as well as statistical methodology related to Bayesian ideas—and that it is time to start exploring how bringing out the inherent Bayesian ways of thinking in education may provide support for learners’ efforts to make sense of the world.

Related science education research emphasizes the role of prior knowledge that students bring with them to the analysis of data and how that prior knowledge impacts the inferences they make about data. For example, \textcite{mkm07} summarized findings across a number of studies showing that when elementary grade students reasoned with data in a context about which they had strong and valid prior understandings that informed their interpretation of data (about force and motion in the context of ramps), and they were able to explain sources of error in the data \parencite{mk03}, but when the data was in a context they did not understand well, they struggled to distinguish between random variation due to sources of error and sources of variation due to something that was causing a difference in the data (generated by measuring the motion of pendulums). Later design- and laboratory-based research has shown that students can develop an understanding of scientific ideas through the analysis of data \parencite{ls04,mkk17}.
